Precision line-by-line forward models for \'echelle spectra

### 2022-08-11 paper/paper1/ms.tex
Change name of the paper for now
add section on how to get to fundamental properties

\begin{document}
\shorttitle{Transfer learning with blas\'e}
\shortauthors{Gully \& Morley}
\title{Automatic line-by-line semi-empirical models for high-grasp stellar and telluric spectra}

@@

\subsection{Line-by-line fundamental parameter estimates}
to-be-completed...

@@

This functional form has a few advantages. It is smooth. The smoothness of the transition is controlled by a tunable parameter, $b$. It can handle either sub- or super-Lorentzian shapes. It is roughly based on the $\chi$-factor\authorcomment1{citation to Hartmann et al. 1989, etc.}. In the limit $\lim_{a\to0} \mathsf{G}$, the lineshape becomes exactly Lorentzian. The sigmoid is efficiently implemented in PyTorch.
Finally, it enforces that the perturbation only produces absorption and not emission profiles.

@@

There exists at least a few demerits of adopting the line lists, and therefore supporting the \emph{blase} strategy.  First, these line lists need to undergo expensive multi-level radiative transfer calculations in order to obtain their amplitudes, so adopting the line lists would mean a laborious and computationally expensive pursuit simply to get close to what has already been computed.  Second, as the effective temperature scales to ultracool dwarfs (Figure \ref{fig_Nlines_vs_teff}) the number of lines sky-rockets, tending towards the billions for T-dwarfs.  The methane line list alone \censorbox{Hargraves et al 2022} represents a prohibitive data volume.  The ExoJAX and Radis \censorbox{Radis paper} libraries offer a breakthrough solution to the voluminous line list problem.  Even still, \emph{blase} deals with the less pure but more practical ``pseudo line'' that gets closer to the astronomical observables anyways, and offers a middle ground between the extremes of interpretability and performance.

>>>>

\begin{document}
\shorttitle{Transfer learning with blas\'e}
\shortauthors{Gully \& Morley}
\title{Precision line-by-line forward models for \'echelle spectra}

@@

\subsection{Line-by-line fundamental parameter estimates}
The conversion of spectral line depths to spectral type\citep{1901AnHar..28..129C}---and by extension $T_\mathrm{eff}$ \citep{1925PhDT.........1P}---has occupied a large chunk of stellar spectroscopy in the last century, and the measurement of fundamental stellar properties remains one of the principal applications of astronomical spectroscopy.  Blas\'e as it currently stands does not output fundamental properties, and in fact it takes them \emph{for granted}: you must specify \emph{e.g.} the $T_\mathrm{eff}$, $\log{g}$, [Fe/H], and [$\alpha$/Fe] of your PHOENIX template as Step Zero before that template gets cloned and warped to match data.  So obtaining a $T_\mathrm{eff}$ estimate is impossible.

Simply put, we seek a function, $f$:
$$f_j(a_j, \sigma_j, \gamma_j) \to (T_\mathrm{eff}, \log{g}, [\mathrm{Fe}/\mathrm{H}])$$

There are $N_{\mathrm{lines}}$ such functions---one for each spectral line---because each line has its own temperature, gravity, and metallicity dependence.  For visualization simplicity purposes, we have assumed the PHOENIX grid is merely 3-dimensional ($T_\mathrm{eff}$, $\log{g}$, [Fe/H]), but the same logic applies to 4-D and higher, and it works for sparse and irregularly sampled grids too.

Fascinatingly, there are two related ways forward towards this function.  They both involved ensembles of spectra, such as a spectral atlas or library, and they both involve first obtaining the inverse function:

\begin{eqnarray}
    g_j(T_\mathrm{eff}, \log{g}, [\mathrm{Fe}/\mathrm{H}]) &\to& (a_j, \sigma_j, \gamma_j)
\end{eqnarray}

A purely model-based approach would start by cloning every single PHOENIX spectrum in the grid dimensions-of-interest, over the wavelength range-of-interest.  The resulting product would be a line-by-line catalog of cloned spectral properties for each 3D coordinate on this grid.  We could then assemble a heatmap of for how each $j^{th}$ line properties changes across this heatmap, $g_j$.  In this way, we are reverse-engineering the \emph{e.g.} temperature dependence of each spectral line, as encoded by the PHOENIX atmosphere models.  The final step would involve finding the neareast neighbor of each $j^{th}$ observed spectral line to each $j^{th}$ grid heatmap point; that is, finding a way to invert the function $g_j$ to get $f_j$.  The information-weighted mean, median, or mode of these nearest neighbors could then be reported as a revised ``best-fit $T_\mathrm{eff}$'', for instance.

A semi-empirical approach could improve on this purely model-based approach.  We know that the PHOENIX line depths, widths, and shapes are imperfect, and so this ($T_\mathrm{eff}$, $\log{g}$, [Fe/H]) heatmap will have large flaws in the line-by-line properties: the contours are systematically too bright or dark.  There are many conceivable ways to quantify these heatmap flaws.  Blas\'e provides an expedient route.  One can pull the heatmap towards the locus of points established by running blas\'e on benchmark systems.  The known $T_\mathrm{eff}$ of such systems would anchor the trend.  We suspect the directionality of the purely model based heatmap must be correct, simply the slope, offset, and concavity of the trends may be wrong. With new information from each benchmark, the heatmap would get lifted like a central tentpole propping up an under-supported tent.

The construction and calibration of these ensembles of spectra represents a tremendous amount of work beyond the scope of this paper.  But its inevitable creation represents a major breakthrough for the field of stellar astrophysics.  It would be an extremely precise, accurate, fast, and interpretable way to measure the properties of stars based solely on their high bandwidth \'echelle spectrum.

\subsection{Abundances}

@@

This functional form has a few advantages. It is smooth. The smoothness of the transition is controlled by a tunable parameter, $b$. It can handle either sub- or super-Lorentzian shapes.  In the limit $\lim_{a\to0} \mathsf{G}$, the lineshape becomes exactly Lorentzian. The sigmoid is efficiently implemented in PyTorch.
Finally, it enforces that the perturbation only produces absorption and not emission profiles.

@@

There exists at least a few demerits of adopting the line lists, and therefore supporting the \emph{blase} strategy.  First, these line lists need to undergo expensive multi-level radiative transfer calculations in order to obtain their amplitudes, so adopting the line lists would mean a laborious and computationally expensive pursuit simply to get close to what has already been computed.  Second, as the effective temperature scales to ultracool dwarfs (Figure \ref{fig_Nlines_vs_teff}) the number of lines sky-rockets, tending towards the billions for T-dwarfs.  The methane line list alone \citep{2020ApJS..247...55H} represents a prohibitive data volume.  The ExoJAX and Radis \cite{2019JQSRT.222...12P,2021JQSRT.26107476V} libraries offer a breakthrough solution to the voluminous line list problem.  Even still, \emph{blase} deals with the less pure but more practical ``pseudo line'' that gets closer to the astronomical observables anyways, and offers a middle ground between the extremes of interpretability and performance.
